---
output: github_document
---
<style>
div.sourceCode {
    overflow: hidden;
}
</style>


<br>
<br>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

> This document serves as a description of contents in `FSI_code.zip` and a guide to reproducing results described in "Over half of western United States' most abundant tree species in decline" (Stanke et al., 2020; `NCOMMS-20-20430`). We describe all system requirements and installation guidelines required to install the publicly available `rFIA` R package, which contains most of the code necessary to reproduce our results (all remaining code available in `FSI_code.zip`). We provide a demonstration of our software (implemented in `rFIA`) as well as specific instructions for software use and replication of results. Contents of `FIA` and `results` directories within `FSI_code.zip` are empty initially. All required data will be downloaded and saved upon running `.R` scripts. 


# System Requirements
We have implemented methods to compute the Forest Stability Index (described in `NCOMMS-20-20430`) using publicly available USDA Forest Inventory and Analysis database in the open source R package, `rFIA`. Methods in the `rFIA` R package and other code used to produce the results outlined in `NCOMMS-20-20430` have been tested on Windows 10, Ubuntu 18.04, and R versions 3.5.1 and 4.0.0. For more on the rFIA R package, check out our [website](https://rfia.netlify.app/).

**The data used to produce results outlined in `NCOMMS-20-20430` may be too large to be processed on a standard desktop computer (RAM limitations may prevent loading/processing in R). We recommend using a machine with at least 64GB of RAM to reproduce our results (all analysis presented in `NCOMMS-20-20430` were completed on a machine with 384GB of RAM).** We recognize that reviewers may not have access to these computational resources. In such event, we have provided an additional alternative workflow that uses a subset of the full dataset for demonstration and review purposes (see lines 51-54 of `estimate.R`).

<br>


# Installation

Users can install the released version of `rFIA` from CRAN with:
``` {r, eval=FALSE}
install.packages('rFIA')
```

or the development version from GitHub with:

``` {r, eval=FALSE}
devtools::install_github('hunter-stanke/rFIA')
```



<br> 

# Software demonstration
As an example, we will implement the Forest Stability Index using USDA Forest Inventory and Analyis (FIA) data collected between 2001-2019 in the state of Colorado, USA. All code shown below can be run in R independent of the contents of `FSI_code.zip`. Processing time will vary with network speed as data must be downloaded from an online repository, however should not exceed 5 minutes on a normal desktop computer. All results will be returned as `data.frame` objects in R.

We first download the FIA data for Colorado using `rFIA`:
```{r, eval = FALSE}
library(rFIA)
library(dplyr)

## Dowloading and loading into R
co <- getFIA('CO')
```

```{r, echo = FALSE}
library(rFIA)
library(dplyr)

## Dowloading and loading into R
co <- readFIA('C:/Users/hstan/Dropbox/FIA/', states = 'CO')
```

We can then use the `fsi` function to estimate the Forest Stability Index for all forest in the state using the 'simple moving average' estimator. As done in `NCOMMS-20-20430`, we allow slopes and intercepts of maximum size density curves to vary by forest type using the `scaleBy` argument:
```{r}
## FSI for all forest
fsi(co, 
    method = 'sma',
    scaleBy = FORTYPCD)
```

where `FSI` is the estimated Forest Stability Index for all forestland in CO by `YEAR`. `SI_STATUS` indicates the status of the population: decline if `FSI` is negative and confidence interval (`FSI_INT`) excludes zero; expand if `FSI` is positive and confidence interval excludes zero; stable otherwise. Estimates can be returned at the plot-level by specifying `byPlot = TRUE`:
```{r}
## FSI plot-level
fsi(co, 
    byPlot = TRUE,
    scaleBy = FORTYPCD) %>%
   ## Forested plots only
   filter(PLOT_STATUS_CD == 1) %>%
   filter(!is.na(PREV_PLT_CN))
```


We can group estimates by species by specifying `bySpecies = TRUE`:
```{r}
## FSI by species
fsi(co, 
    method = 'sma', 
    bySpecies = TRUE,
    scaleBy = FORTYPCD)
```


or by any other grouping variable contained in the `co` database using the `grpBy` argument:
```{r}
## Grouping by ownership class
fsi(co, 
    method = 'sma',
    grpBy = OWNGRPCD,
    scaleBy = FORTYPCD)
```




# Instructions for use
## _Applying the FSI to our study region_
To apply the `FSI` to our study region (i.e., western US), we simply expand our population of interest to include Washington, Oregon, California, Idaho, Montana, Utah, Nevada, Arizona, New Mexico, and Colorado: 

```{r eval=FALSE}
## Download and load data for all states
db <- getFIA(c('WA', 'OR', 'CA', 'NV', 'AZ',
               'NM', 'CO', 'UT', 'MT', 'ID'))

## Estimate the FSI by species across the full
## study region (range-wide indices)
rangeWide <- fsi(db, 
                 method = 'sma',
                 bySpecies = TRUE,
                 scaleBy = FORTYPCD)
```


These data are large (~8GB), and download speeds will depend on a users network connection. Once downloaded, processing should not exceed 10 minutes. 


_Reproducing of our results_

To reproduce the results outlined in `NCOMMS-20-20430`, modify working directory in `estimate.R`, `climate.R`, and `model.R` to the location where `FSI_code` is unzipped. Then simply source the `.R` files listed at the root of the `FSI_code` directory with the following (must be run in this sequence): 

```{r eval=FALSE}
## Download data and estimate the FSI by 
## populations of interest
source('./FSI_code/estimate.R')


## Run disturbance model
source('./FSI_code/disturbMod.R')
```


The above will (1) download all FIA data required for the analysis, and save the data in the `FIA` directory. (2) Produce estimates of the FSI for populations of interest (i.e., species). Results will be stored in the `results` directory, with plot-level estimates found in `results/plt`, ecoregion-scale estimates found in `results/spatial` and range-wide estimates found in `results/regionWide`. Model results (estimated coefficients and associated credible intervals) will be stored in `results/model`.

